### 2012-01-25 docs/notes.tex
Update TODO.  Add notes and eckmann analysis.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%% ICML 2012 EXAMPLE LATEX SUBMISSION FILE %%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% Use the following line _only_ if you're still using LaTeX 2.09.
%\documentstyle[icml2012,epsf,natbib]{article}
% If you rely on Latex2e packages, like most moden people use this:
\documentclass{article}

% For figures
\usepackage{graphicx} % more modern
%\usepackage{epsfig} % less modern
\usepackage{subfigure}

% For citations
\usepackage{natbib}

% For algorithms
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{verbatim}
\usepackage{url}
\usepackage{graphicx, color}
\usepackage{amsmath}

% As of 2011, we use the hyperref package to produce hyperlinks in the
% resulting PDF.  If this breaks your system, please commend out the
% following usepackage line and replace \usepackage{icml2012} with
% \usepackage[nohyperref]{icml2012} above.
\usepackage{hyperref}

% Packages hyperref and algorithmic misbehave sometimes.  We can fix
% this with the following command.
\newcommand{\theHalgorithm}{\arabic{algorithm}}

% Employ the following version of the ``usepackage'' statement for
% submitting the draft version of the paper for review.  This will set
% the note in the first column to ``Under review.  Do not distribute.''
\usepackage{icml2012} 
% Employ this version of the ``usepackage'' statement after the paper has
% been accepted, when creating the final version.  This will set the
% note in the first column to ``Appearing in''
% \usepackage[accepted]{icml2012}

% The \icmltitle you define below is probably too long as a header.
% Therefore, a short form for the running title is supplied here:
\icmltitlerunning{Relational Event Blockmodels}

@@

\twocolumn[
\icmltitle{Relational Event Blockmodels}

% It is OKAY to include author information, even for blind
% submissions: the style file will automatically remove it for you
% unless you've provided the [accepted] option to the icml2012
% package.
\icmlauthor{Your Name}{email@yourdomain.edu}
\icmladdress{Your Fantastic Institute,
            314159 Pi St., Palo Alto, CA 94306 USA}
\icmlauthor{Your CoAuthor's Name}{email@coauthordomain.edu}
\icmladdress{Their Fantastic Institute,
            27182 Exp St., Toronto, ON M6H 2T1 CANADA}

% You may provide any keywords that you 
% find helpful for describing your paper; these are used to populate 
% the "keywords" metadata in the PDF but will not be shown in the document
\icmlkeywords{boring formatting information, machine learning, ICML}

\vskip 0.3in
]

\begin{abstract} 
ICML 2012 full paper submissions are due February 24, 2012. Reviewing will
be blind to the identities of the authors, and therefore identifying
information must not appear in any way in papers submitted for review. Submissions must be in PDF, 8 page length limit.
\end{abstract}

\pagestyle{empty}

@@

Borrowing from the intuition of stochastic blockmodels, we propose a hierarchical model of continuous time network data that learns latent clusters of nodes which share similar patterns of interaction.  [TODO: Explain how our experiments show that this helps.]  By combining a latent variable framework with a local dependence model for event sequences, we provide a hierarchical approach for modeling event-based network data, accompanying recent hierarchical extensions for latent position models \cite{Handcock2007} and exponential random graph models \cite{Schweinberger2011}.
\begin{figure*}
\center
 \def\svgwidth{5.5in}
  \input{example1.pdf_tex}
\caption{Illustration of event data and the assumptions of the model.  Left: An sequence of two events among four nodes: (1,2) occurs at time $t_1$ and (2,3) occurs at time $t_2$.  Center: The intensity functions $\lambda_{1,2}(t)$ (solid) and $\lambda_{34}(t)$ (dotted).  Note $\lambda_{34}$ can only change when events occur involving either node 3 or node 4.  Right: Illustration of the block model, where events in cluster 1 (circles) are governed by the vector $\beta_{1,1}$, events in cluster 2 (squares) are governed by $\beta_{2,2}$, and an circle-to-square event governed by $\beta_{1,2}$.}
\label{fig:example}
\end{figure*}  
\section{Model}

@@

In general, we may have different types of events that can occur.   In the context of relational events occurring among $N$ nodes, for example, we let the \emph{risk set} $\mathcal{R}$ include all allowed dyads.  Let each dyad $(i,j) \in \mathcal{R}$ have an intensity function $\lambda_{ij}(t|\cdot)$ that is piecewise constant.  As shown in Figure \ref{fig:example}, we assume each intensity function $\lambda_{ij}(t)$ may only change following an event involving either $i$ or $j$.  Let $\tau_{i,j,m}$ be the time that $\lambda_{i,j}$ changed before the $m$th event and let $\mathcal{R}_{i,j}$ be the set of dyads involving either $i$ or $j$.  % Letting $ \mathcal{R}_{k_1,k_2}$ be the set of events that can occur from $k_1$ to $k_2$ and  $\boldsymbol{\tau}_{i,j}$ is the set of knots where the respective intensities may change, we may write the likelihood as follows:
% \begin{align}
% \mathcal{L}(\mathcal{A}|\theta) %&= \prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot) \prod_{(i,j) \in \mathcal{R}} \prod_{k=1}^{|\boldsymbol{\tau}_{ij}|} \exp\{ - (\tau_{i,j,k} - \tau_{i,j,k-1}) \lambda_{ij}(\tau_{i,j,k} | \cdot) \} \\
% &= \prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot) \prod_{k_1=1}^K \prod_{k_2=1}^K \prod_{(i,j) \in \mathcal{R}_{k_1,k_2}} \prod_{r=1}^{|\boldsymbol{\tau}_{i,j}|} \exp\{ - (\tau_{k_1,k_2,r} - \tau_{k_1,k_2,r-1}) \lambda_{ij}(\tau_{i,j,r} | \cdot) \} 
% \end{align}
% \begin{align}
% \mathcal{L}(\mathcal{A}|\theta) &= \prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot) \prod_{r=1}^N \exp\{ - (t_m - \tau_{i,r,m}) \lambda_{ir}(t_m | z_i,z_r)  - (t_m - \tau_{r,j,m}) \lambda_{rj}(t_m | z_r,z_j) \} 
% \end{align}
\begin{align}
\mathcal{L}(\mathcal{A}|\theta) =& \prod_{m=1}^M 
\lambda_{i_m,j_m}(t_m|\cdot) \times \\
&\prod_{(i,j) \in \mathcal{R}_{i_m,j_m}}\exp\{ - (t_m - \tau_{i,j,m})
\lambda_{ij}(t_m | z_i,z_j) \}
\label{eqn:llk}
\end{align}

@@

@@

To perform Gibbs sampling for a given cluster assignment $z_i$, we need to compute the likelihood for each possible assignment of $z_i$. 
This operation can be done in parallel, and only the portion of the likelihood involving the newly assigned cluster  need to be recomputed.  Computing the likelihood is then $O(T \cdot P \cdot 4N \cdot K)$.

@@

\section{Simulation}
\begin{figure*}
\center
\includegraphics[width=1.75in]{../figs/syn-mat.pdf}
\includegraphics[width=3in]{../figs/syn-counts.pdf}
\includegraphics[width=1.5in]{../figs/syn-llk.pdf}
\caption{Illustration of 1000 simulated events, as described in text. Left: Counts of each dyad. Center: Boxplot of distribution of participation counts across dyads.  The top left shows an increased propensity for reciprocity within cluster 1; bottom right shows more AB-AY events within cluster 2. Right: Loglikelihood vs. iteration during MCMC.}
\label{fig:syncounts}
\end{figure*}

@@

TODO: Compare to fitting the relational event model without the blockmodel structure.  Compare to relational event blockmodel without covariates: $$p(y_{ij}) = \exp\{\beta_{z_i,z_j}\} / \sum_{(a,b) \in \mathcal{R}}\exp\{\beta_{z_{a},z_{b}}\} $$
Or more simply, just use $s_{ij}(t) = 1$.

@@

% TODO: Discuss possible variations on the prior structure of $\beta$: 1. Enforce $\beta_{k,l} = \beta_{l,k}$. 2. Assume $\beta_{k,l,p} \sim \mbox{Normal}(\beta_p,\sigma_p^2)$. 3. Place other priors that provide different kinds of regularization.

% TODO: Discuss decision to condition on $z$ for covariate vector $s_{ij}$.

% TODO: Discuss choice of $\log \lambda$ instead of just $\lambda$.

% TODO: Discuss choice against time-varying effects.

% TODO: Discuss if the semiparametric partial likelihood approach can be used in this framework.

% \begin{table}[t]
% \caption{Classification accuracies for naive Bayes and flexible 
% Bayes on various data sets.}
% \label{sample-table}
% \vskip 0.15in
% \begin{center}
% \begin{small}
% \begin{sc}
% \begin{tabular}{lcccr}
% \hline
% \abovespace\belowspace
% Data set & Naive & Flexible & Better? \\
% \hline
% \abovespace
% Breast    & 95.9$\pm$ 0.2& 96.7$\pm$ 0.2& $\surd$ \\
% Cleveland & 83.3$\pm$ 0.6& 80.0$\pm$ 0.6& $\times$\\
% Glass2    & 61.9$\pm$ 1.4& 83.8$\pm$ 0.7& $\surd$ \\
% Credit    & 74.8$\pm$ 0.5& 78.3$\pm$ 0.6&         \\
% Horse     & 73.3$\pm$ 0.9& 69.7$\pm$ 1.0& $\times$\\
% Meta      & 67.1$\pm$ 0.6& 76.5$\pm$ 0.5& $\surd$ \\
% Pima      & 75.1$\pm$ 0.6& 73.9$\pm$ 0.5&         \\
% \belowspace
% Vehicle   & 44.9$\pm$ 0.6& 61.5$\pm$ 0.4& $\surd$ \\
% \hline
% \end{tabular}
% \end{sc}
% \end{small}
% \end{center}
% \vskip -0.1in
% \end{table}

% Acknowledgements should only appear in the accepted version. 
\section*{Acknowledgments}

% In the unusual situation where you want a paper to appear in the
% references without citing it in the main text, use \nocite
\appendix
\section{Simulation}

@@

\bibliography{refs}
\bibliographystyle{icml2012}

@@

% This document was modified from the file originally made available by
% Pat Langley and Andrea Danyluk for ICML-2K. This version was
% created by Lise Getoor and Tobias Scheffer, it was slightly modified  
% from the 2010 version by Thorsten Joachims & Johannes Fuernkranz, 
% slightly modified from the 2009 version by Kiri Wagstaff and 
% Sam Roweis's 2008 version, which is slightly modified from 
% Prasad Tadepalli's 2007 version which is a lightly 
% changed version of the previous year's version by Andrew Moore, 
% which was in turn edited from those of Kristian Kersting and 
% Codrina Lauth. Alex Smola contributed to the algorithmic style files.

>>>>

\pagestyle{empty}
\documentclass[11pt]{article}
\usepackage{graphicx}
\usepackage{verbatim}
\usepackage{url}
\usepackage{graphicx, color}
\usepackage{amsmath}
%\usepackage[small,compact]{titlesec} 
%\usepackage[small,it]{caption}

\textwidth 6.5in
\headheight .1in
\topmargin -.2 in
\textheight 8.7in
\oddsidemargin 0in
\evensidemargin 1in

@@

 \centerline{\Large \bf Blockmodels for Dynamic Network Data} 
 \medskip
\centerline{\bf Christopher DuBois}
 \bigskip

@@

Borrowing from the intuition of stochastic blockmodels, we propose a hierarchical model of continuous time network data that learns latent clusters of nodes which share similar patterns of interaction.  [TODO: Explain how our experiments show that this helps.]  By combining a latent variable framework with a local dependence model for event sequences, we provide a hierarchical approach for modeling event-based network data, accompanying recent hierarchical extensions for latent position models \cite{Handcock2007} and exponential random graph models \cite{Schweinberger2011}.

\section{Model}

@@

In general, we may have different types of events that can occur.   In the context of relational events occurring among $N$ nodes, for example, we let the \emph{risk set} $\mathcal{R}$ include all allowed dyads.  Let each dyad $(i,j) \in \mathcal{R}$ have an intensity function $\lambda_{ij}(t|\cdot)$ that is piecewise constant.  As shown in Figure \ref{fig:example}, we assume each intensity function $\lambda_{ij}(t)$ may only change following an event involving either $i$ or $j$.  Let $\tau_{i,j,m}$ be the time that $\lambda_{i,j}$ changed before the $m$th event and let $\mathcal{R}_{i,j}$ be the set of dyads involving either $i$ or $j$.  % Letting $ \mathcal{R}_{k_1,k_2}$ be the set of events that can occur from $k_1$ to $k_2$ and  $\boldsymbol{\tau}_{i,j}$ is the set of knots where the respective intensities may change, we may write the likelihood as follows:
% \begin{align}
% \mathcal{L}(\mathcal{A}|\theta) %&= \prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot) \prod_{(i,j) \in \mathcal{R}} \prod_{k=1}^{|\boldsymbol{\tau}_{ij}|} \exp\{ - (\tau_{i,j,k} - \tau_{i,j,k-1}) \lambda_{ij}(\tau_{i,j,k} | \cdot) \} \\
% &= \prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot) \prod_{k_1=1}^K \prod_{k_2=1}^K \prod_{(i,j) \in \mathcal{R}_{k_1,k_2}} \prod_{r=1}^{|\boldsymbol{\tau}_{i,j}|} \exp\{ - (\tau_{k_1,k_2,r} - \tau_{k_1,k_2,r-1}) \lambda_{ij}(\tau_{i,j,r} | \cdot) \} 
% \end{align}
% \begin{align}
% \mathcal{L}(\mathcal{A}|\theta) &= \prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot) \prod_{r=1}^N \exp\{ - (t_m - \tau_{i,r,m}) \lambda_{ir}(t_m | z_i,z_r)  - (t_m - \tau_{r,j,m}) \lambda_{rj}(t_m | z_r,z_j) \} 
% \end{align}
\begin{align}
\mathcal{L}(\mathcal{A}|\theta) &= \prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot) \prod_{(i,j) \in \mathcal{R}_{i_m,j_m}}\exp\{ - (t_m - \tau_{i,j,m}) \lambda_{ij}(t_m | z_i,z_j) \}
\end{align}

\begin{figure}
 \def\svgwidth{6in}
  \input{example1.pdf_tex}
\caption{Illustration of event data and the assumptions of the model.  Left: An sequence of two events among four nodes: (1,2) occurs at time $t_1$ and (2,3) occurs at time $t_2$.  Center: The intensity functions $\lambda_{1,2}(t)$ (solid) and $\lambda_{34}(t)$ (dotted).  Note $\lambda_{34}$ can only change when events occur involving either node 3 or node 4.  Right: Illustration of the block model, where events in cluster 1 (circles) are governed by the vector $\beta_{1,1}$, events in cluster 2 (squares) are governed by $\beta_{2,2}$, and an circle-to-square event governed by $\beta_{1,2}$.}
\label{fig:example}
%\includegraphics[width=6in]{example}
\end{figure}

@@

\subsection{Submodels}

The above framework generalizes several important special cases.  The first is one where events within each block share the rate at which they occur.  This can be implemented by only having an intercept statistic, $s_0(t,i,j) = 1$.  As these intensities do not change under this specification, the likelihood simplifies to 
$$\mathcal{L}(\mathcal{A}|\beta) = \prod_{m=1}^M \lambda_{i_m,j_m} \prod_{(i,j) \in \mathcal{R}} \exp\{-t_M \lambda_{i,j}\}$$
where $ \lambda_{ij} = \exp\{\boldsymbol{\beta}_{z_i,z_j}\}$.  Each block in this model is homogeneous Poisson process where the dyad associated with each event is draw uniformly from the dyads in that block.

Another important model is one where within group interactions are governed via a set of parameters while all between group interactions are governed by a different set.  In this case $\boldsymbol{\beta}_{k1,k2} = \boldsymbol{\gamma}$ for all $k1 \ne k2$.

@@

To perform Gibbs sampling for a given cluster assignment $z_i$, we need to compute the likelihood for each possible assignment of $z_i$. 
This operation can be done in parallel, and only the portion of the likelihood involving the newly assigned cluster  need to be recomputed.  Computing the likelihood is then $O(T \cdot P \cdot 4N)$, though we essentially split this computation into $K^2$ pieces.

@@

\section{Simulation}
\begin{figure}
\center
%\includegraphics[width=1.75in]{../figs/syn-mat.pdf}
%\includegraphics[width=3in]{../figs/syn-counts.pdf}
%\includegraphics[width=1.5in]{../figs/syn-llk.pdf}
\caption{Illustration of 1000 simulated events, as described in text. Left: Counts of each dyad. Center: Boxplot of distribution of participation counts across dyads.  The top left shows an increased propensity for reciprocity within cluster 1; bottom right shows more AB-AY events within cluster 2. Right: Loglikelihood vs. iteration during MCMC.}
\label{fig:syncounts}
\end{figure}

@@

TODO: Compare to fitting the relational event model without the blockmodel structure.  Compare to relational event blockmodel without covariates: $p(y_{ij}) = \exp\{\beta_{z_i,z_j}\} / \sum_{(a,b) \in \mathcal{R}}\exp\{\beta_{z_{a},z_{b}}\} $
Or more simply, just use $s_{ij}(t) = 1$.

@@

TODO: Discuss possible variations on the prior structure of $\beta$: 1. Enforce $\beta_{k,l} = \beta_{l,k}$. 2. Assume $\beta_{k,l,p} \sim \mbox{Normal}(\beta_p,\sigma_p^2)$. 3. Place other priors that provide different kinds of regularization.

TODO: Discuss if the semiparametric partial likelihood approach can be used in this framework.

\appendix
\section{Simulation}

@@

\bibliographystyle{plain}
\bibliography{refs}

@@
